Modelling complex population structure using $F$-statistics and Principal Component Analysis

### 2021-07-12 fstats_pca_lewontin.tex
better discussion

\subsection{Introduction to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2 &=& \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos{\phi}\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

@@

\begin{align}
F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2  &&\nonumber\\ 
&=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
&=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}\label{eq:fpc}
\end{align}
A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}.
As $F_3$ and $F_4$ can be written as sums of $F_2$-terms (Eqs. \ref{eq:f3fromf2}, \ref{eq:f4fromf2}), analogous relations apply.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to create all figures and analyses will be available on [Archive to be created before publishing].

@@

\paragraph{Computation $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} \url{https://github.com/uqrmaie1/admixtools} to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.

@@

The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. 
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$. $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B).

@@

 Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot. In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics. This is because they are outside the $n$-ball in higher dimensions. The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0.

@@

\section{Discussion}
Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using specific models using a lower number of populations. It could also allows for calculation of $F$-statistics involving unsampled populations, which can be useful for checking assumptions.

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. 
However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008}. This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow. In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA. There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.

@@

\appendix
\section{Derivation}\label{appendix:fonpc}
\begin{eqnarray}
F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
&=& \sum_k (P_{ik} - P_{jk})^2
\end{eqnarray}

@@

Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic. As these are scalars, they do not change above calculation.
\bibliographystyle{plainnat}
\bibliography{main}

>>>>

\subsection{Introduction to $F$-statistics}
PCA is typically used to model population structure between many populations. $F$-statistics take the opposite approach, revealing the relationship between just two,  three or four populations at a time. The three $F$-statistics can be defined as 
\begin{subequations}
	\begin{align}
	&F_2(X_1, X_2) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})^2
	&=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_2 \rangle = \frac{1}{S}\normsq{X_1-  X_2}\\
	&F_3(X_1; X_2, X_3) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{1l} - x_{3l}) &=& \frac{1}{S}\langle X_1 - X_2, X_1 - X_3 \rangle\\	
	&F_4(X_1, X_2; X_3, X_4) &=& \frac{1}{S}\sum_{l=1}^S(x_{1l} - x_{2l})(x_{3l} - x_{4l}) &=& \frac{1}{S}\langle X_1 - X_2, X_3 - X_4 \rangle	\text{,}
	\end{align}
\end{subequations}
where $\norm{\cdot}$ denotes the Euclidean norm and $\langle \cdot, \cdot \rangle$ denotes the dot product. The normalization by the number of SNPs $S$ is assumed to be the same for all calculations and is thus omitted subsequently. Some elementary properties of the dot product between vectors $a, b, c$ that I will use later are
\begin{subequations}
\begin{align}
\langle a, b \rangle &= \sum_i a_ib_i\\
\langle a, b \rangle &= \norm{a}\norm{b}\cos(\phi)\\
\langle a, a \rangle &= \normsq{a}\\
\langle a + c, b \rangle &= \langle a, b \rangle + \langle b, c \rangle,
\end{align}
\end{subequations}
where $\phi$ is the angle between $a$ and $b$.

@@

\begin{align}
F_2(X_i, X_j) &=&  \sum_{l=1}^L \big( x_{il} -x_{jl}\big)^2  &&\nonumber\\ 
&=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2   &=& F_2(Y_i, Y_j) \nonumber\\
&=& \sum_{k=1}^n (p_{ik} - p_{jk})^2  &=& F_2(P_i, P_j) \text{,}\label{eq:fpc}
\end{align}
A derivation of this change-of-basis is given in Appendix \ref{appendix:fonpc}, Equation \ref{eq:changeofbasis}.
As $F_3$ and $F_4$ can be written as sums of $F_2$-terms (Eqs. \ref{eq:f3fromf2}, \ref{eq:f4fromf2}), analogous relations apply.

@@

\section{Material \& Methods}
The theory outlined in the previous section suggests that $F$-statistics have a geometric interpretation in PCA-space, which can be approximated on PCA plots. In the next section I explore this connection in detail, and illustrate it on two sample data sets that I briefly introduce here. Both are based on the analyses by \cite{lazaridis2014}. The data is from the Reich lab compendium data set (v44.3), downloaded from \url{https://reich.hms.harvard.edu/allen-ancient-dna-resource-aadr-downloadable-genotypes-present-day-and-ancient-dna-data}, using data on the ``Human Origins''-SNP set (597,573 SNPs). SNPs with missing data in any population are excluded. The code used to create all figures and analyses will be available on \url{https://github.com/BenjaminPeter/fstats_pca}.

@@

\paragraph{Computing $F$-statistics and PCA}
I perform analyses at the level of populations to ease presentation.  It is an assumption of $F$-statistics that the genetic variation within sampled population is independent of the variation between samples \citep{patterson2012}. All computations are performed in \texttt{R}. I use \texttt{admixtools 2.0.0} (\url{https://github.com/uqrmaie1/admixtools}) to compute $F$-statistics. To obtain a PC-decomposition, I first calculate all pairwise $F_2$-statistics, obtain a nearby negative semidefinite matrix using the \texttt{nearPD} function, and then use equation \ref{eq:mds} and the \texttt{eigen} function to obtain the PCs.

@@

The $F_3$-statistic gives a more precise interpretation: we are looking for the space where $F_3$ is negative, i.e. 
\begin{eqnarray}
2 F_3(X_x; X_1, X_2) &=& 2\langle  X_x - X_1, X_x - X_2 \rangle \nonumber\\
      &=& \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} < 0 \label{eq:f3neg}
\end{eqnarray}
By the Pythagorean theorem, $F_3 = 0 $ if and only if $X_1, X_2$ and $X_x$ form a right-angled triangle. The associated region where $F_3=0$ is a $n$-sphere (or a circle in two dimensions) with diameter $\overline{X_1X_2}$. $F_3$ is negative when the triangle is obtuse, i.e. $X_x$ is admixed if it lies inside the $n$-ball with diameter $\overline{X_1X_2}$ (Figure \ref{fig:geom}B, Equation \ref{eq:f3sphere}).

@@

 Thus if $\hat{F}_3 \ll F_3$, the true circle will be bigger than would be predicted from a 2D-plot. In this case, substantial relevant genetic differentiation is ``hidden'' in the higher PCs, and populations that appear inside the circle on a PCA-plot may, in fact, have positive $F_3$-statistics. This is because they are outside the $n$-ball in higher dimensions. The converse interpretation is more strict: if a population lies outside the circle on \emph{any} 2D-projection, $F_3$ is guaranteed to be bigger than 0 (see Equation \ref{eq:f3circle} in the Appendix).

@@

\section{Discussion}
Particularly for the analysis of ancient DNA, $F$-statistics  are  a powerful tool to describe population genetic diversity. Here, I show that the geometry of $F$-statistics \citep{oteo-garcia2021} leads to a number of simple interpretations of $F$-statistics on a PCA plot. This allows for direct and quantitative comparisons between $F$-statistic-based results and PCA biplots. As PCA is often ran in an early step in data analysis, this also aids in generation of hypotheses that can be more directly evaluated using generative models, (e.g using a lower number of populations). It also allows reconciling apparent contradictions between $F$-statistics and PCA-plots; differences between the two data summaries are explained solely by higher PCs, and so whenever such contradictions arise, and so higher PCs will be informative for population structure. Previous interpretation of PCA in the context of population genetic models have primarily focused on the PCs, which can be derived analytically for trees \citep{cavalli-sforza1975} and homogeneous spatial models \citep{novembre2008a}. My interpretation here is different in that it puts more emphasis on the geometry itself, rather than directly interpreting the PCs. One consequence is that the results here are not impacted by sample ascertainment and sample sizes \citep{mcvean2009, novembre2008a}, which are common concerns in the interpretation of PCA. However, a very skewed sampling distribution will increases the likelihood that more or different PCs will have to be included in the analysis. From this perspective, one could envision a framework where $F$-statistics are used to decide which samples should be included to obtain a low-dimensional PCA-plot ``representative'' of the data

As $F$-statistics are motivated by trees, they assume that populations are discrete, related as a graph, and that gene flow between populations is rare \citep{patterson2012,harney2021}. However, in many regions, all humans populations are admixed to some degree \citep{pickrell2014}, and in regions such as Europe, genetic diversity is distributed continuously \citep{novembre2008, novembre2008a}. This provides a challenge for interpretation; as many $F_3$ and $F_4$ statistics may indicate gene flow. In my example (Figure \ref{fig:f3}A), most Southern European populations are ``admixed'' between Basques and Turkish, but a more accurate model might be one of continuous variation where Basque and Turkish lie on one of multiple gradients; which is more directly visualized with PCA. There are a number of tools that have been developed that use multiple $F$-statistics to build complex models, such as \texttt{qpGraph} \citep{lazaridis2014} and \texttt{qpAdm} \citep{harney2021}. One issue with these approaches is that they are usually restricted to at most a few dozen populations. As ancient DNA data sets now commonly include thousands of individuals, analysts are faced with the challenge of which data to include. A common approach is to sample a large number of distinct models, and retain the ones that are compatible with the data. However, as both \texttt{qpGraph} and \texttt{qpAdm} assume that gene flow is rare and discrete, selecting sets of populations that did experience little gene flow will provide good fits. One example of this is the world foci data set used here, which contains only 33 populations from across the world, and which is well-approximated by two PCs. However, this ascertainment misses a large amount of variation; a more dense sampling would show that in many places human genetic diversity is very gradual and multi-layered \citep{lazaridis2014, peter2020a}. The PCA-based  interpretation offers an alternative that trades interpretability for robustness. Particularly interpreting a (normalized) $F_4$-statistic as a correlation coefficient translates to generalized models of gene flow. Separating $F$-statistics in a sum of model and residuals, and performing a PCA on the latter (such as in Figure \ref{fig:f4}D)  is another way how we can visualize $F$-statistics and evaluate the model fit.

@@

\appendix
\section{Derivations}\label{appendix:fonpc}
\setcounter{equation}{0}
\renewcommand{\theequation}{\thesection\arabic{equation}}
Depending on a readers' background in linear algebra, these results may appear elementary; I include them here for reference and because they were not obvious to me at the onset of this project.
\paragraph{$F$-statistics are invariant under a change-of-basis}
\begin{eqnarray}
F_2(X_i, X_j) &=& \sum_{l=1}^L \big( (x_{il} - \mu_l) -(x_{jl} -\mu_l)\big)^2 = F_2(Y_i, Y_j)\nonumber\\
&=& \sum_{l=1}^L \big( \sum_k L_{kl}P_{ik} - \sum_kL_{kl}P_{kj}\big)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl} (P_{ik} -P_{jk}) \right)^2\nonumber\\
&=& \sum_{l=1}^L \left( \sum_k L_{kl}^2 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'} L_{kl}L_{k'l}(P_{ik} - P_{jk'})^2 \right)\nonumber\\
&=& \sum_k \underbrace{\left(\sum_{l=1}^L L_{kl}^2\right)}_1 (P_{ik} -P_{jk})^2 + 2\sum_{k\neq k'}\underbrace{\left(\sum_{l=1}^L L_{kl}L_{k'l}\right)}_{0} (P_{ik} - P_{jk'})^2\nonumber\\
&=& \sum_k (P_{ik} - P_{jk})^2 \label{eq:changeofbasis}
\end{eqnarray}

@@

Note that if we estimate $F_2$, unbiased estimators are obtained by subtracting the population-heterozygosities $H_i, H_j$ from the statistic. As these are scalars, they do not change above calculation.

\paragraph{The region of negative $F_3$-statistics is a $n$-ball}
Without loss of generality, assume that $X_1 = (r, 0, 0, \dots)$ and $X_2 = (-r, 0, 0, \dots)$, and let us assume that $X_x$ has coordinates $(x_1, x_2, \dots, x_S)$ Assuming $F_3(X_x; X_1, X_2) = 0$, equation \ref{eq:f3neg} becomes
\begin{align}
2 F_3(X_x; X_1, X_2) &=  \normsq{X_x - X_1} + \normsq{X_x - X_2}  - \normsq{X_1 - X_2} = 0 \nonumber\\
&= \left[(x_1-r)^2 + \sum_{i=2}^S x_i^2\right] + \left[(x_1 + r)^2 + \sum_{i=2}^S x_i^2\right] - 4r^2 \nonumber\\
&= 2 \left[\sum_{i=1}^S x_i^2 + r^2 + x_1r - x_1r\right] - 4r^2\nonumber\\
F_3(X_x; X_1, X_2) & = -r^2 + \sum_{i=1}^S x_i^2 = -r^2 + \normsq{X_x} = 0  \text{,} \label{eq:f3sphere}
\end{align}
which is the equation of a $n$-sphere with radius $r$ and center at the origin, as assumed from the placing of $X_1$ and $X_2$. Now, assume that $F_3$ is negative, i.e. $F_3(X_x; X_1, X_2) = -k < 0$. Moving $r^2$ to the left we obtain
\begin{equation}
r^2 - k = \normsq{X_x} \text{,}
\end{equation}
which is another $n$-sphere with a smaller radius, showing that all points inside the $n$-sphere will have negative $F_3$-values.

\paragraph{If a population lies outside the circle of this $n$-Sphere in any 2D-projection, $F_3$ is positive} 
Assume the center of the $n$-sphere $C = \frac{X_1 + X_2}{2} = (c_1, c_2, \dots c_S)$, and $X_x = (x_1, x_2, \dots x_S)$.  Then, 
\begin{align}
F_3(X_x; X_1, X_2) &= \normsq{X_x - C} - r^2 \nonumber\\
&= \underbrace{(x_1 - c_1)^2 + (x_2 - c_2)^2}_{>r^2} + \underbrace{\sum_{i=3}^S (x_i - c_i)^2}_{\geq 0} - r^2 \nonumber\\
& > 0 \text{.}\label{eq:f3circle}
\end{align}
The condition $(x_1 - c_1)^2 + (x_2 - c_2)^2 >r^2$ is satisfied whenever $X_x$ is outside the circle obtained from projecting the $n$-sphere on the first two dimensions. An analogous argument applies for any low-dimensional representation.

\bibliographystyle{plainnat}
\bibliography{main}
